GEE models

### 2017-07-20 doc/manuscript.Rmd
add bit about refering to esm methods for covariate selection
updated the conclusions section in light of reviewer comments

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC). See
the ESM Methods for more detail on QIC and the directed acyclic graphs. 
`r cite_st('qic')` shows the covariates compared using QIC. We selected the
model (M9) that had the fewest covariates and that had similar fit between the
outcome measures. The final GEE model included the covariates years from baseline
(time), sex (female or male), ethnicity (European ancestry or non-European 
ancestry), baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, 
and baseline age were time-*independent* variables. After scaling, 
and log-transforming the model variables, then exponentiating the GEE estimates,
the interpretation of the GEE results is as an expected percent difference in 
the outcome variable for every standard deviation (SD) increase in the predictor
variable given the covariates are held constant. With the main effect 
associations (not with a time interaction), this interpretation is applied 
consistently at each time point and thus represents associations of baseline 
NEFA for each future value of the outcomes. Since TAG is a risk factor for
diabetes and since NEFA contribute to TAG production [@Barrows2006a], TAG may
act as a mediator between NEFA and the outcomes. To determine the role of TAG in
the association between NEFA and the outcomes, TAG was included in the GEE model
in a sensitivity analysis. Lastly for the GEE models, we tested for an 
interaction with sex, ethnicity, or time by the predictor term for each outcome 
variable.

@@

In a Canadian population of adults who were at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta-cell
function after 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty
acids as a proportion, an observation that was further confirmed by multivariate
cluster techniques, which found no ability of these
individual NEFA to predict beta-cell function or dysglycemia. These observations suggest
that the absolute size of the total NEFA fraction, rather than its specific
composition, likely influences the pathogenesis of diabetes, at least within a
population at-risk for diabetes.

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures over time. However, we believe this
is a strength for our specific objective, as the chance of reverse causality is 
reduced given that fatty acid and glucose metabolism pathways are tightly 
integrated. This is also an observational cohort, and there may be some residual
confounding we have not considered or that could not be measured. Nonetheless,
potential covariates were empirically analyzed prior to inclusion into the
GEE models to best understand and minimize potential
confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female and European-ancestry and as such our results may not be
generalizable to other populations. However, given these limitations, our study
also has several strengths, including the longitudinal design and the rigorous
statistical techniques and methods applied, which are specifically suited to
investigating temporal relationships and to handling the multivariate nature of
the data. Lastly, our cohort contains highly detailed and comprehensive
variable measurements at each collection visit, and has both concentration and
mol% data for the fatty acids.

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta-cell function
[@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta-cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta-cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta-cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta-cell function. However, in modeling these fatty acids as a mol% and
using novel clustering analysis approaches, no specific fatty acids in the NEFA
fraction strongly predicted lower beta-cell function, insulin sensitivity, or
dysglycemia status. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific
composition of fatty acids, that is responsible for the hypothesized lipotoxic
effects of NEFA on the beta-cells.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw no association of any individual or total NEFA with hepatic
(HOMA2-%S) or whole-body insulin sensitivity (ISI). We identified
WC as having a powerful attenuating effect on the associations between
unadjusted and adjusted models. There are some possible explanations for our
null findings for the insulin sensitivity measures and the influence of WC on
the results. First, WC may be a strong causal link in previously reported
associations between NEFA and insulin sensitivity, given the role adipocytes
play in metabolism (e.g. leptin, adiponectin). Second, there may be differences
in physiology between fasting and postprandial NEFA kinetics that we are not
able to investigate but that may explain our null findings for fasting NEFA and
insulin sensitivity. For instance, some experimental studies using clamp
protocols found that fasting NEFA was a weak predictor of insulin sensitivity
compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue (e.g. as seen in obesity and larger WC) following postprandial TAG
lipolysis via lipoprotein lipase may result in NEFA spillover into the blood and
a subsequent increase in circulating NEFA [@Almandoz2013a], which may be more
metabolically active given postprandial activity. Third, the null findings seen
may be due to the high risk population examined in PROMISE, as the majority of subjects had a 
high WC and BMI. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat more advanced
stage in the pathogenesis of diabetes.

@@

In conclusion, we found that total NEFA was a strong predictor for lower
beta-cell function over 6 years, irrespective of the specific composition of the
NEFA fraction, suggesting that efforts at lowering circulating total NEFA
through medication and/or lifestyle strategies to reduce the risk of diabetes
are well warranted. While future
studies are needed to confirm these findings, our results reinforce the
importance of continuing to investigate the role of circulating NEFA
concentration on the natural history of diabetes.

>>>>

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC). See
the ESM Methods for more detail on QIC and the directed acyclic graphs. 
`r cite_st('qic')` shows the covariates compared using QIC. We selected the
model (M9) that had the fewest covariates and that had similar fit between the
outcome measures. The final GEE model included the covariates years from baseline
(time), sex (female or male), ethnicity (European ancestry or non-European 
ancestry), baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, 
and baseline age were time-*independent* variables. See the ESM Methods for more
detail on the covariate inclusion. After scaling,
and log-transforming the model variables, then exponentiating the GEE estimates,
the interpretation of the GEE results is as an expected percent difference in 
the outcome variable for every standard deviation (SD) increase in the predictor
variable given the covariates are held constant. With the main effect 
associations (not with a time interaction), this interpretation is applied 
consistently at each time point and thus represents associations of baseline 
NEFA for each future value of the outcomes. Since TAG is a risk factor for
diabetes and since NEFA contribute to TAG production [@Barrows2006a], TAG may
act as a mediator between NEFA and the outcomes. To determine the role of TAG in
the association between NEFA and the outcomes, TAG was included in the GEE model
in a sensitivity analysis. Lastly for the GEE models, we tested for an 
interaction with sex, ethnicity, or time by the predictor term for each outcome 
variable.

@@

In a Canadian population of adults who were at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta-cell
function after 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty acids as a
proportion, an observation that was further confirmed by multivariate cluster
techniques, which found no ability of these individual NEFA to predict beta-cell
function. We did find that one fatty acid as a mol%, eicosapentaenoic acid
(20:5n-3), had a positive association with insulin sensitivity. In general,
these observations suggest that the absolute size of the total NEFA fraction,
rather than its specific composition (aside from EPA), likely influences the 
pathogenesis of diabetes, at least within a population at-risk for diabetes.

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures over time. However, we
believe this is a strength for our specific objective, as the potential impact
that reverse causality, due to the tight integration between fatty acid and
glucose metabolism, may have on the results is lower. This is also an
observational cohort, and there may be some residual confounding we have not
considered or that could not be measured. Nonetheless, potential covariates were
empirically analyzed prior to inclusion into the GEE models to best understand
and minimize potential confounding. While there may be potential overadjustment
due to controlling for WC, both QIC and directed acyclic graph outputs
recommended including WC in the GEE model. Even still, we ran sensitivity
analyses and identified WC as the covariate that attenuated the unadjusted
results. This attenuation may be due to the fact that higher NEFA may be 
reflective of greater adipose tissue mass and acting as a mediator between the 
adipose tissue and insulin sensitivity (since adipocytes are active endocrine 
organs).

Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female and European-ancestry and, while sex and ethnicity did
not influence the results presented from this cohort, our findings may not
entirely be generalizable to other populations. However, given these
limitations, our study also has several strengths, including the longitudinal
design and the rigorous statistical techniques and methods applied, which are
specifically suited to investigating temporal relationships and to handling the
multivariate nature of the data. Lastly, our cohort contains highly detailed and
comprehensive variable measurements at each collection visit, and has both
concentration and mol% data for the fatty acids.

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta-cell function
[@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta-cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta-cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta-cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta-cell function. No fatty acid as a mol% had associations with
beta-cell function. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific 
composition of fatty acids, that is responsible for the hypothesized lipotoxic 
effects of NEFA on the beta-cells.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw in general no association of any individual or total NEFA with
hepatic (HOMA2-%S) or whole-body insulin sensitivity (ISI), except for EPA. 
There is some evidence [@Steffen2015a;@Virtanen2014;@Sarbolouki2013] that EPA
may be involved with lowering the risk for diabetes, potentially due to its role
in inflammation [@Calder2009]. However, given we found the association only with
HOMA2-%S and not with ISI, the significant association could be chance finding,
which will require future studies to confirm. We also identified WC as having a
powerful attenuating effect on the associations between unadjusted and adjusted
models. There are some possible explanations for our null findings for the
insulin sensitivity measures and the influence of WC on the results. First, WC
may be a strong causal link in previously reported associations between NEFA and
insulin sensitivity, given the role adipocytes play in metabolism (e.g. leptin,
adiponectin). Second, there may be differences in physiology between fasting and
postprandial NEFA kinetics that we are not able to investigate but that may
explain our null findings for fasting NEFA and insulin sensitivity. For
instance, some experimental studies using clamp protocols found that fasting
NEFA was a weak predictor of insulin sensitivity compared to postprandial
concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
which may be more metabolically active given postprandial activity. Third, the
null findings seen may be due to the high risk population examined in PROMISE, 
as the majority of subjects had a high WC and BMI. It may be that in this
population, insulin resistance has become well established and NEFA may not
contribute to insulin resistance at this somewhat more advanced stage in the
pathogenesis of diabetes.

@@

In conclusion, we found that total NEFA was a strong predictor for lower 
beta-cell function over 6 years, irrespective of the specific composition of the
NEFA fraction, suggesting that efforts at lowering circulating total NEFA 
through medication and/or lifestyle strategies to reduce the risk of diabetes 
are well warranted. We also found that higher EPA was associated with higher
hepatic insulin sensitivity. While future studies are needed to confirm these
findings, our results reinforce the importance of continuing to investigate the
role of circulating NEFA concentration on the natural history of diabetes.
